---
title: "Individual Assingment 5 JM"
author: June McNally
date: 2026-05-10
format: pdf
execute:
  warning: false
  message: false
---

# Set Up

## Packages

```{r}
#| label: packages
# geberal and cleaning
library(tidyverse)
library(here)
library(janitor)
library(snakecase)

# wrapping axis labels
library(scales)

# reading in .xlsx files
library(readxl)

# visualizing missing data
library(naniar)

# arranging plots
library(patchwork)

```

## Data

```{r}
#| label: data

# taxonomic information
taxon_list <- read_csv(here("data", "taxon_list.csv")) 

# aquatic invertebrates
aq_ins <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                    sheet = "Aquatic Insects") 

# site information
sites <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                   sheet = "sites")

# water quality information
water_quality <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                           sheet = "Water Quality",
                           na = c("", "n/a", "N/A", "over", "173+"))

```

## Class code
 
```{r}
#| label: ncos-site-object

# creating new object from sites
ncos_sites <- sites |> 
  # filter to only include NCOS sites sampled four times a year
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")
```

```{r}
#| label: cleaning-taxon-list

# creating new clean object from taxon_list
taxon_list_clean <- taxon_list |> 
  # converting taxon name to snake case (no spaces or capital letters,
  # only underscores)
  mutate(verbatim_name = to_any_case(verbatim_name, case = "snake"))
```

```{r}
#| label: water-quality-cleaning

# creating new clean object from water quality
water_quality_clean <- water_quality |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites in ncos_sites
  semi_join(ncos_sites, by = "site") |> 
  # create new column to join with later data frames
  unite("date_site", date, site, remove = TRUE) |> 
  # group by date_site column
  group_by(date_site) |> 
  # calculate median pH, salinity, DO
  summarize(med_ph = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # filter out salinity outliers
  filter(date_site != "2022-08-12_NVBR") 
```

```{r}
#| label: cleaning-aq-ins

# creating new clean object from aquatic inverts
aq_ins_clean <- aq_ins |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites occurring in ncos_sites
  semi_join(ncos_sites, by = c("site")) |> 
  # renaming columns
  rename(date = date_on_vial) |> 
  # select columns of interest
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  # filter to only include "filtered beaker" samples
  filter(sample_type %in% c("FB 250", "FB250")) |> 
  # convert the data frame to long format
  pivot_longer(cols = ostracod:annelida_polychaete,
               names_to = "taxon",
               values_to = "abundance") |> 
  # group by date, site, taxon
  group_by(date, site, taxon) |> 
  # calculate average abundance per liter (sum abundance divided by 7.5 liters)
  summarize(ave_lit = sum(abundance, na.rm = TRUE)/7.5) |> 
  # ungroup data frame
  ungroup() |> 
  # create a new column called `date_site` to join with water quality
  unite("date_site", date, site, remove = FALSE) |> 
  # join with water quality
  left_join(water_quality_clean, by = c("date_site")) |> 
  # join with taxon list
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  # add full names of sites
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "South of Venoco Bridge",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~  "West Pond",
    site == "NDC" ~ "Devereux Creek"
  )) |> 
  # setting factor levels for site
  mutate(site_full = fct_reorder(site_full, med_sal, .fun = "median", .na_rm = TRUE),
         site_full = fct_rev(site_full))
```

```{r}
#| label: copepod-cleaning

# creating new object from clean aquatic inverts 
copepods <- aq_ins_clean |> 
  # filter to only include copepods
  filter(taxon == "copepod") |> 
  # log + 1 transform mean abundance per liter
  mutate(log_ave_lit = log(ave_lit + 1)) 
```

```{r}
#| label: visualizing-copepods

copepods_plot <- ggplot(data = copepods,
                   mapping = aes(x = site_full, y = log_ave_lit)) +
  geom_jitter(height = 0, shape = 21, width = 0.1) +
  stat_summary(geom = "point", fun = "median", size = 3, color = "blue") +
  scale_x_discrete(labels = label_wrap(15)) +
  labs(x = "Site", y = "log + 1 transformed average abundance/L", title = "Copepods") +
  theme_bw()

copepods_plot
```

```{r}
#| label: visualizing-salinity
#| fig-width: 8
#| fig-height: 6

salinity <- ggplot(data = aq_ins_clean,
       mapping = aes(x = site_full,
                     y = med_sal)) +
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21) +
  scale_x_discrete(labels = label_wrap(width = 15)) +
  stat_summary(fun = median,
               color = "red",
               size = 0.5) +
  labs(x = "Site",
       y = "Median salinity (ppt)",
       title = "Site salinity") +
  theme_bw()

salinity
```

 
# Problems

## 1. Visualizing differences across sites in major taxon abundance

### a. Planning figure

-X axis: site (site_full)
-Y axis: ostracod abundance (log + 1 transform)
-geometry to represent average abundance/L: geom_jitter()
-geometry to represent medians: stat_summary()

### b. Wrangle data

```{r}
ostracods <- aq_ins_clean |> 
  filter(taxon == "ostracod") |> 
  mutate(log_ave_lit = log(ave_lit + 1))

slice_sample(ostracods, n = 10)
```

### c. Coding figure

```{r}
#| label: visualizing-ostracods
#| fig-width: 10
#| fig-height: 6

ostracod_plot <- ggplot(data = ostracods,
                   mapping = aes(x = site_full,
                                 y = log_ave_lit)) +
  # small open circles for individual measurements
  geom_jitter(height = 0,
              shape = 21,
              width = 0.1) +
  # blue filled point for median
  stat_summary(geom = "point",
               fun = "median",
               size = 3,
               color = "blue") +
  # wrap long site names
  scale_x_discrete(labels = label_wrap(15)) +
  # labels
  labs(x = "Site",
       y = "log + 1 transformed average abundance/L",
       title = "Ostracods") +
  # same theme as key
  theme_bw()

ostracod_plot
```

### d. Adding ostracod panel

```{r}
#| label: multipanel-figure
#| fig-width: 10
#| fig-height: 14

salinity / copepods_plot / ostracod_plot
```

### e. Interpretation of the figure

The abundance of copepods north and south of Venoco Bridge and at Phelps Road is higher than the abundance of ostracods, the abundance of each is similar in the Main Channel, south of Phelps Bridge, north of Phelps Bridge, and at Devereux creek. There does not seem to be a strong correlation between of the copepods and ostracods with the salinity.









